---
title: "Reproducible Research Assignment 2"
author: "Daren Thomas"
date: "5 May 2016"
output: html_document
---

## Synopsis

## Data Processing

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can [download the file from the course web site](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2): 

```{r cache=TRUE} 
noaa.url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
noaa.bz2 = "storm-data.csv.bz2"
if (! file.exists(noaa.bz2)) {
    download.file(url=noaa.url, destfile=noaa.bz2)
}
noaa <- read.csv(noaa.bz2)
```

After reading, we check the first few rows (there are `r nrow(noaa)` ) in the dataset:

```{r}
dim(noaa)
```

```{r}
head(noaa)
```

Also, I am interested in the types of the fields:

```{r}
str(noaa)
```

It looks like the variables I'm most interested in are:

- `EVTYPE`
- `FATALITIES`
- `INJURIES`
- `PROPDMG`
- `CROPDMB`

### Processing of Field `EVTYPE`
```{r}
library(dplyr)
```

```{r}
nrow(distinct(select(noaa, EVTYPE)))
```

The field `EVTYPE` has `r nrow(distinct(select(noaa, EVTYPE)))` distinct values. Going through them, it seems some are summaries. Others have labels that start with spaces. There are some inconsistencies with uppercase/lowercase - let's just move everything to uppercase. I am also going to check for typos.

First, remove surrounding whitespace on event types and uppercase everything:

```{r}
noaa$EVTYPE2 <- trimws(noaa$EVTYPE)
noaa$EVTYPE2 <- toupper(noaa$EVTYPE2)
nrow(distinct(select(noaa, EVTYPE2)))
```

The new field `EVTYPE2` has `r nrow(distinct(select(noaa, EVTYPE2)))`, less than `EVTYPE`, so we are making progress.

Next, remove any summary rows, and while we are at it, any NA rows (those with value `?`):

```{r}
noaa2 <- filter(noaa, !grepl('summary', noaa$EVTYPE2, ignore.case=TRUE))
noaa2 <- filter(noaa2, EVTYPE2 != '?')
nrow(distinct(select(noaa2, EVTYPE2)))
```

The new dataset `noaa2` has only `nrow(distinct(select(noaa2, EVTYPE2)))` distinct event types.

Sorting the field has the advantage that it makes it easy to spot typos in the data (such as `AVALANCE` and `AVALANCHE`)

```{r}

# NOTE: the code below will generate a long list of event types
evtypes <- distinct(select(noaa2, EVTYPE2)) %>% arrange(EVTYPE2)
head(evtypes)
```

Next, we slog through this list, but we're lazy, so we are going to compare using `adist` and assume that a distance < 2 is a typo. We will update the original event type with the match that has the most entries, assuming that the typo is less common than the original value. This does create a problem of `THUNDERSTORM WIND 50` being lumped togeter with `THUNDERSTORM WIND 60`. So, for this analysis, I am going to remove all numbers from event types as well as numbers that start with a letter (`F0`, `G50` etc.) and parenthesis too.

```{r}
# remove numbers with potential leading letter from event types
noaa2$EVTYPE3 <- gsub('[A-Z]?[0-9]+', '', noaa2$EVTYPE2)

# remove ' MPH'
noaa2$EVTYPE3 <- sub(' MPH', '', noaa2$EVTYPE3)

# remove any non-letter character from event types
noaa2$EVTYPE3 <- gsub('[^A-Z]', '', noaa2$EVTYPE3)
```

Next we update similar event types, using the most common one as the probably correct one:
(NOTE: I didn't just create the whole matrix/vectorize, because R Studio kept crashing... therefore we loop!)
(NOTE: this process is iterative, do it a couple of times...)

```{r}
evtypes <- distinct(select(noaa2, EVTYPE3)) %>% arrange(EVTYPE3)
for (row in 1:(nrow(evtypes)-1)) {
    left <- evtypes[row, 1]
    right <- evtypes[row+1, 1]
    distance <- adist(left, right)[1, 1]
    if (distance < 2) {
        nleft <- length(noaa2[noaa2$EVTYPE3 == left, 'EVTYPE3'])
        nright <- length(noaa2[noaa2$EVTYPE3 == right, 'EVTYPE3'])
        if (nleft > nright) {
            noaa2$EVTYPE3[noaa2$EVTYPE3 == right] <- left
            #print(c(right, length(noaa2[noaa2$EVTYPE3 == right, 'EVTYPE3'])))
        }
        else {
            noaa2$EVTYPE3[noaa2$EVTYPE3 == left] <- right
            #print(c(left, length(noaa2[noaa2$EVTYPE3 == left, 'EVTYPE3'])))
        }
    }
}
evtypes <- distinct(select(noaa2, EVTYPE3)) %>% arrange(EVTYPE3)
```

It turns out that this approach is only a first stab - we'd have to actually define a smaller set of event types and a mapping to those types. For the purposes of this analysis, I think we have done enough.

Let us look at the other fields we'd like to analyse:

```{r}
noaa2$FATALITIES <- as.numeric(noaa2$FATALITIES)
noaa2$INJURIES <- as.numeric(noaa2$INJURIES)
noaa2$PROPDMG <- as.numeric(noaa2$PROPDMG)
noaa2$CROPDMG <- as.numeric(noaa2$CROPDMG)
```

Going through the documentation, it seems as if the `PROPDMGEXP` and `CROPDMGEXP` fields contain an exponend to multiply the `PROPDMG` and `CROPDMG` values with. I need to clean this up since a couple of conventions are used:

```{r}
# first all uppercase
noaa2$PROPDMGEXP <- toupper(noaa2$PROPDMGEXP)
noaa2$PROPDMGEXP <- trimws(noaa2$PROPDMGEXP)

# replace `?` and `-` and `+` and ` ` with `0`
noaa2$PROPDMGEXP[noaa2$PROPDMGEXP %in% c('?', '-', '', '+')] <- '0'

# replace K, M, B, H with 3, 6, 9, 2 respectively
noaa2$PROPDMGEXP[noaa2$PROPDMGEXP == 'K'] <- '3'
noaa2$PROPDMGEXP[noaa2$PROPDMGEXP == 'M'] <- '6'
noaa2$PROPDMGEXP[noaa2$PROPDMGEXP == 'B'] <- '9'
noaa2$PROPDMGEXP[noaa2$PROPDMGEXP == 'H'] <- '2'
noaa2$PROPDMGEXP <- as.numeric(noaa2$PROPDMGEXP)


# create new field with multiplied values
mutate(noaa2, property.damage = PROPDMG * (10 ^ PROPDMGEXP))

# repeat for crop damage
# first all uppercase
noaa2$PROPDMGEXP <- toupper(noaa2$PROPDMGEXP)
noaa2$PROPDMGEXP <- trimws(noaa2$PROPDMGEXP)

# replace `?` and `-` and `+` and ` ` with `0`
noaa2$CROPDMGEXP[noaa2$CROPDMGEXP %in% c('?', '-', '', '+')] <- '0'

# replace K, M, B, H with 3, 6, 9, 2 respectively
noaa2$CROPDMGEXP[noaa2$CROPDMGEXP == 'K'] <- '3'
noaa2$CROPDMGEXP[noaa2$CROPDMGEXP == 'M'] <- '6'
noaa2$CROPDMGEXP[noaa2$CROPDMGEXP == 'B'] <- '9'
noaa2$CROPDMGEXP[noaa2$CROPDMGEXP == 'H'] <- '2'
noaa2$CROPDMGEXP <- as.numeric(noaa2$CROPDMGEXP)


# create new field with multiplied values
mutate(noaa2, crop.damage = CROPDMG * (10 ^ CROPDMGEXP))
```

## Analysis

I am trying to answer these two questions:

1. Across the United States, which types of events (as indicated in the ð™´ðš…ðšƒðšˆð™¿ð™´ variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?




## Results

